---
title: "index"
format: html
editor: visual
---

#TODO: Should I keep this as Qmd or a base R file is enough -- I will have more clarity as I move forward making changes 

# Overall Goal


Our goal is to write functions that will manipulate and process data sets that come from a census API. We’ll
create generic functions to automatically summarize and plot certain returned data

# Data

We are going to query the Public Use Microdata Sample (PUMS) Census API. This API gives pseudo
person-level data

## PUMS API URLs

We aren’t going to worry about Puerto Rico data so focusing on the basic PUMS information

- [2021](https://api.census.gov/data/2021/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24&key=YOUR_KEY_GOES_HERE)
- [2022](https://api.census.gov/data/2022/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24&key=YOUR_KEY_GOES_HERE)
- [2023](https://api.census.gov/data/2023/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24&key=YOUR_KEY_GOES_HERE)
- [2024](https://api.census.gov/data/2024/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24&key=YOUR_KEY_GOES_HERE)

## List of PUMS variables 

- [2021 list of variables](https://api.census.gov/data/2021/acs/acs1/pums/variables.html)
- [2022 list of variables](https://api.census.gov/data/2022/acs/acs1/pums/variables.html)
- [2023 list of variables](https://api.census.gov/data/2023/acs/acs1/pums/variables.html)
- [2024 list of variables](https://api.census.gov/data/2024/acs/acs1/pums/variables.html)

## Users choice

Allow user to choose certian variables and conditions 
TODO: come back to this and fill what is allowed 

# Part 1: Data Processing 

## Parsing Variable Values

We’re going to start by downloading and parsing the JSON files with the variable descriptions for each year for the following years
- 2021
- 2022
- 2023
- 2024

```{r}
library(httr)
library(tidyverse)
library(dotenv)

# load in env file for API key

variables_base_list <- list( "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX")
variables_2021_2022 <- append(variables_base_list,list("ST","STATE"))
variables_2023_2024 <- append(variables_base_list,list("REGION","DIVISION"))


variables_by_year <- list(
  "2021" = variables_2021_2022,
  "2022" = variables_2021_2022,
  "2023" = variables_2023_2024,
  "2024" = variables_2023_2024
)

get_variables_by_year <- function(year_str){
  
  if(!is.character(year_str)) {
    warning("Year should be passed as character string like \"2024\". coercing to character string to make code work")
    year_str <- as.character(year_str)
  }
  
  url <- paste0(
    "https://api.census.gov/data/",
    year_str,
    "/acs/acs1/pumspr/variables.json"
  )
  api_variables <- GET(url)
  
  print(status_code(api_variables))
  
  print(names(api_variables))          # does this include "variables" at all?
  print(length(api_variables$variables))
  
   variables_response <- fromJSON(rawToChar(api_variables$content), simplifyVector = TRUE)

 # variables_response$variables
 # names(variables_response$variables)
 # variables_by_year[year_str]
 filtered_variables <-   variables_response$variables[names(variables_response$variables) %in% variables_by_year[[year_str]]]
  
}

# for (year in names(variables_by_year)) {
#   var_yr_variables_temp <- get_variables_by_year(year)
# 
#   cat("Year:", year,
#       "- length(names):", length(names(var_yr_variables_temp)),
#       "- length(sapply):", length(sapply(var_yr_variables_temp, function(v) v$predicateType)),
#       "\n")
# }

# var_yr_variables_temp <- get_variables_by_year("2024")
# variable_table_2024 <- data.frame(
#   variable = names(var_yr_variables_temp),
#   predicateType = sapply(var_yr_variables_temp, function(v) v$predicateType)
# )
# 
# print(variable_table_2024)
# 
# var_yr_variables_temp <- get_variables_by_year("2023")
# variable_table_2023 <- data.frame(
#   variable = names(var_yr_variables_temp),
#   predicateType = sapply(var_yr_variables_temp, function(v) v$predicateType)
# )
# 
# print(variable_table_2023)



var_yr_variables_temp <- get_variables_by_year("2022")

print(status_code(GET(paste0("https://api.census.gov/data/2022/acs/acs1/pumspr/variables.json"))))
cat("length(names):", length(names(var_yr_variables_temp)),
    "- length(sapply):", length(sapply(var_yr_variables_temp, function(v) v$predicateType)), "\n")

# var_yr_variables_temp <- get_variables_by_year("2021")
# variable_table_2021 <- data.frame(
#   variable = names(var_yr_variables_temp),
#   predicateType = sapply(var_yr_variables_temp, function(v) v$predicateType)
# )
# 
# print(variable_table_2021)


for (year in names(variables_by_year)) {
  
  file_name <- paste0("data/variables_", year, ".rds")
  
  saveRDS(
    get_variables_by_year(year),
    file_name
  )
}


```




## Parsing API Data

Write a function to query the PUMS API and return a nicely formatted tibble

- 2021
- 2022
- 2023
- 2024

```{r}

library(jsonlite)
library(tidyverse)
library(dotenv)

# load in env file for API key
load_dot_env()

api_key <- Sys.getenv("CENSUS_API_KEY")

load_variables_rds_by_year <- function(year_str  = c("2024", "2023", "2022", "2021")) {
  year_str <- match.arg(year_str, several.ok = FALSE)
  file_name <- paste0("data/variables_", year_str, ".rds")
  readRDS(file_name)
}



get_pums_data_as_tibble__by_year <- function(api_key, year_str = c("2024", "2023", "2022", "2021"), numeric_vars = c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP")) {
  
   #check for missing inputs 
  if (missing(api_key)) {
    stop("apiKey are required")
  }
  
   year_str <- match.arg(year_str, several.ok = FALSE)
  
  if (missing(year_str)) {
    message("year_str is not supplied, 2024 is supplied as default")
  }
  variables_supported = load_variables_rds_by_year(year_str)
   numeric_vars = intersect(
   numeric_vars,
    names(load_variables_rds_by_year(year_str))
  )
  
  url <- paste0(
    "https://api.census.gov/data/",
    year_str,
    "/acs/acs1/pums?get=",
     paste(numeric_vars, collapse = ","),
     "&SCHL=24&key=",
    api_key
  )
  

  
  pums_data <- GET(url)
  
  resp_json <- fromJSON(rawToChar(pums_data$content))

  resp_tibble <- as_tibble(resp_json)
  
}


pums_tibble_2024 <- get_pums_data_as_tibble__by_year(api_key, c("2024"), numeric_vars = c("AGEP", "PWGTP"))

print(pums_tibble_2024)



```


