---
title: "index"
format: html
editor: visual
  markdown: 
    wrap: 72
---

#TODO: Should I keep this as Qmd or a base R file is enough -- I will have more clarity as I move forward making changes

# Overall Goal

Our goal is to write functions that will manipulate and process data sets that come from a census API. We’ll create generic functions to automatically summarize and plot certain returned data

# Data

We are going to query the Public Use Microdata Sample (PUMS) Census API. This API gives pseudo person-level data

## PUMS API URLs

We aren’t going to worry about Puerto Rico data so focusing on the basic PUMS information

- [2021](https://api.census.gov/data/2021/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24&key=YOUR_KEY_GOES_HERE)
- [2022](https://api.census.gov/data/2022/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24&key=YOUR_KEY_GOES_HERE)
- [2023](https://api.census.gov/data/2023/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24&key=YOUR_KEY_GOES_HERE)
- [2024](https://api.census.gov/data/2024/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24&key=YOUR_KEY_GOES_HERE)

## List of PUMS variables

- [2021 list of variables](https://api.census.gov/data/2021/acs/acs1/pums/variables.html)
- [2022 list of variables](https://api.census.gov/data/2022/acs/acs1/pums/variables.html)
- [2023 list of variables](https://api.census.gov/data/2023/acs/acs1/pums/variables.html)
- [2024 list of variables](https://api.census.gov/data/2024/acs/acs1/pums/variables.html)

## Users choice

Allow user to choose certian variables and conditions TODO: come back to this and fill what is allowed

# Part 1: Data Processing

## Parsing Variable Values

We’re going to start by downloading and parsing the JSON files with the variable descriptions for each year for the following years - 2021 - 2022 - 2023 - 2024

```{r}
library(httr)
library(tidyverse)
library(dotenv)

# load in env file for API key

variables_base_list <- list( "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX")
variables_2021_2022 <- append(variables_base_list,list("ST","REGION","DIVISION"))
variables_2023_2024 <- append(variables_base_list,list("STATE","REGION","DIVISION"))


variables_by_year <- list(
  "2021" = variables_2021_2022,
  "2022" = variables_2021_2022,
  "2023" = variables_2023_2024,
  "2024" = variables_2023_2024
)

get_variables_by_year <- function(year_str){
  
  if(!is.character(year_str)) {
    warning("Year should be passed as character string like \"2024\". coercing to character string to make code work")
    year_str <- as.character(year_str)
  }
  
  url <- paste0(
    "https://api.census.gov/data/",
    year_str,
    "/acs/acs1/pumspr/variables.json"
  )
  api_variables <- GET(url)
  
  print(status_code(api_variables))
  
  print(names(api_variables))          # does this include "variables" at all?
  print(length(api_variables$variables))
  
   variables_response <- fromJSON(rawToChar(api_variables$content), simplifyVector = TRUE)

 # variables_response$variables
 # names(variables_response$variables)
 # variables_by_year[year_str]
 filtered_variables <-   variables_response$variables[names(variables_response$variables) %in% variables_by_year[[year_str]]]
  
}

# for (year in names(variables_by_year)) {
#   var_yr_variables_temp <- get_variables_by_year(year)
# 
#   cat("Year:", year,
#       "- length(names):", length(names(var_yr_variables_temp)),
#       "- length(sapply):", length(sapply(var_yr_variables_temp, function(v) v$predicateType)),
#       "\n")
# }

# var_yr_variables_temp <- get_variables_by_year("2024")
# variable_table_2024 <- data.frame(
#   variable = names(var_yr_variables_temp),
#   predicateType = sapply(var_yr_variables_temp, function(v) v$predicateType)
# )
# 
# print(variable_table_2024)
# 
# var_yr_variables_temp <- get_variables_by_year("2023")
# variable_table_2023 <- data.frame(
#   variable = names(var_yr_variables_temp),
#   predicateType = sapply(var_yr_variables_temp, function(v) v$predicateType)
# )
# 
# print(variable_table_2023)



var_yr_variables_temp <- get_variables_by_year("2022")

print(status_code(GET(paste0("https://api.census.gov/data/2022/acs/acs1/pumspr/variables.json"))))
cat("length(names):", length(names(var_yr_variables_temp)),
    "- length(sapply):", length(sapply(var_yr_variables_temp, function(v) v$predicateType)), "\n")

# var_yr_variables_temp <- get_variables_by_year("2021")
# variable_table_2021 <- data.frame(
#   variable = names(var_yr_variables_temp),
#   predicateType = sapply(var_yr_variables_temp, function(v) v$predicateType)
# )
# 
# print(variable_table_2021)


for (year in names(variables_by_year)) {
  
  file_name <- paste0("data/variables_", year, ".rds")
  
  saveRDS(
    get_variables_by_year(year),
    file_name
  )
}



```

## Parsing API Data

Write a function to query the PUMS API and return a nicely formatted tibble

- 2021
- 2022
- 2023
- 2024

```{r}

library(jsonlite)
library(tidyverse)
library(dotenv)

# load in env file for API key
load_dot_env()

api_key <- Sys.getenv("CENSUS_API_KEY")

load_variables_rds_by_year <- function(year_str  = c("2024", "2023", "2022", "2021")) {
  year_str <- match.arg(year_str, several.ok = FALSE)
  file_name <- paste0("data/variables_", year_str, ".rds")
  readRDS(file_name)
}

```



```{r}

get_pums_data_as_tibble__by_year <- function(api_key,
                                             year_str = c("2024", "2023", "2022", "2021"),
                                             numeric_vars = c("AGEP", "PWGTP")) {
  #check for missing inputs
  if (missing(api_key)) {
    stop("apiKey are required")
  }
  
  year_str <- match.arg(year_str, several.ok = FALSE)
  
  if (missing(year_str)) {
    message("year_str is not supplied, 2024 is supplied as default")
  }
  variables_supported = load_variables_rds_by_year(year_str)
  
  allowed_numeric_vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
  non_allowed_nuermic_var_in_input <- FALSE
  is_pwgtp_part_of_input <- FALSE
  number_of_numeric_values_outside_pwgtp <- 0
  
  # validate the input tags
  for (v in numeric_vars) {
    if (v == "PWGTP") {
      is_pwgtp_part_of_input <- TRUE
    }
    v_allowed <- FALSE
    for (allowed in allowed_numeric_vars) {
      if (v == allowed) {
        v_allowed <- TRUE
      }
    }
    
    if (!v_allowed) {
      non_allowed_nuermic_var_in_input <- TRUE
    }
    else if (v != "PWGTP") {
      number_of_numeric_values_outside_pwgtp <- number_of_numeric_values_outside_pwgtp + 1
    }
  }
  
  if (non_allowed_nuermic_var_in_input) {
    stop("numeric_vars must only contain: ",
         paste(allowed_numeric_vars, collapse = ", "))
  }
  
  if (number_of_numeric_values_outside_pwgtp == 0) {
    stop("at least one numeric var other than PWGTP must be specified in the input")
  }
  
  if (!is_pwgtp_part_of_input) {
    numeric_vars <- c(numeric_vars, "PWGTP")
  }
  
  url <- paste0(
    "https://api.census.gov/data/",
    year_str,
    "/acs/acs1/pums?get=",
    paste(numeric_vars, collapse = ","),
    "&SCHL=24&key=",
    api_key
  )
  
  pums_data <- GET(url)
  resp_json <- fromJSON(rawToChar(pums_data$content))
  #resp_tibble <- as_tibble(resp_json)
  
  pums_tibble <- as_tibble(resp_json[-1, , drop = FALSE]) # without -1 header becomes the first row 
  colnames(pums_tibble) <- resp_json[1, ]
  pums_tibble
}

```


```{r}
# testing
pums_tibble_2024 <- get_pums_data_as_tibble__by_year(api_key, c("2024"), numeric_vars = c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP"))
print(pums_tibble_2024)

#pums_tibble_2023 <- get_pums_data_as_tibble__by_year(api_key, c("2023"), numeric_vars = c("AGEP"))
#print(pums_tibble_2023)

#pums_tibble_2022 <- get_pums_data_as_tibble__by_year(api_key, c("2022"))
#print(pums_tibble_2022)

#pums_tibble_2021 <- get_pums_data_as_tibble__by_year(api_key, c("2021"))
#print(pums_tibble_2021)

##pums_tibble_2024_default <- get_pums_data_as_tibble__by_year(api_key)
#print(pums_tibble_2024_default)

```



```{r}

library(lubridate)

parse_time_label_midpoint <- function(label) {
  if(is.null(label) || is.na(label)) {
    return(as.POSIXct(NA))
  }
  
  parts <- strsplit(label, " to ")[[1]]
  
  if(length(parts) != 2) {
    return(as.POSIXct(NA))
  }
  
  start_clean <- gsub("\\.","",parts[1]) #double esacping it for R syntax 
  end_clean   <- gsub("\\.","",parts[2])
  
  start_time <- lubridate::parse_date_time(start_clean, orders = "IMp")
  end_time <- lubridate::parse_date_time(end_clean, orders = "IMp")
  
  if(is.na(start_time) || is.na(end_time)) {
    return(as.POSIXct(NA))
  }
  
  start_time + (end_time - start_time) / 2
}


convert_time_codes_to_midpoints <- function(codes, variable_metadata) {
  result <- sapply(codes, function(code) {
    if(is.na(code) || code == "") {
      return(as.POSIXct(NA))
    }
    
    label <- variable_metadata$values$item[[code]]
    if(is.null(label)) {
      return(as.POSIXct(NA))
    }
    parse_time_label_midpoint(label)
  })
   as.POSIXct(result, origin = "1970-01-01")
}

result <- parse_time_label_midpoint("6:30 a.m. to 6:34 a.m.")
print(result)
str(result)
class(result)

```



```{r}

get_pums_data_as_tibble__by_year_formatted <- function(api_key,
                                                       year_str = c("2024", "2023", "2022", "2021"),
                                                       numeric_vars = c("AGEP", "PWGTP")) {
  pums_tibble <- get_pums_data_as_tibble__by_year(api_key, year_str, numeric_vars)
  
  variables_metadata <- get_variables_by_year(year_str)
  
  for (col_name in colnames(pums_tibble)) {
    #print(variables_metadata[col_name])
    if (col_name %in% c("JWAP", "JWDP")) {
      pums_tibble[[col_name]] <- convert_time_codes_to_midpoints(pums_tibble[[col_name]], variables_metadata[[col_name]])
    } else if (col_name %in% numeric_vars) {
      pums_tibble[[col_name]] <- as.numeric(pums_tibble[[col_name]])
      # } else if (col_name %in% categorical_vars) {
      #   pums_tibble[[col_name]] <- factor(pums_tibble[[col_name]], ...)
    }
  }
  pums_tibble
}

pums_tibble_2024_formatted <- get_pums_data_as_tibble__by_year_formatted(
  api_key,
  c("2024"),
  numeric_vars = c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
)
print(pums_tibble_2024_formatted)


``` 

```{r}
match_allowed_value <- function(value, allowed_values) {
   for (allowed in allowed_values) {
      if (toupper(value) == toupper(allowed)) {
        return(allowed)
      }
   }
     return(NA)
}
```

```{r}
get_pums_data_as_tibble__by_year_formatted_with_categorical_variables <-  function(api_key,
                                                                                   year_str = c("2024"),
                                                                                   numeric_vars = c("AGEP", "PWGTP"),
                                                                                   categorical_vars = c("SEX"),
                                                                                   geography_level = c("state"),
                                                                                   geography_value = c("37")) {
  #check for missing inputs
  if (missing(api_key)) {
    stop("apiKey are required")
  }
  
  year_str <- match.arg(year_str, several.ok = FALSE)
  
  if (missing(year_str)) {
    message("year_str is not supplied, 2024 is supplied as default")
  }
  variables_supported <- load_variables_rds_by_year(year_str)
  
  allowed_numeric_vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
  non_allowed_nuermic_var_in_input <- FALSE
  is_pwgtp_part_of_input <- FALSE
  number_of_numeric_values_outside_pwgtp <- 0
  
  # validate the input tags
  for (v in numeric_vars) {
    if (v == "PWGTP") {
      is_pwgtp_part_of_input <- TRUE
    }
    v_allowed <- FALSE
    for (allowed in allowed_numeric_vars) {
      if (v == allowed) {
        v_allowed <- TRUE
      }
    }
    
    if (!v_allowed) {
      print(v)
      non_allowed_nuermic_var_in_input <- TRUE
    }
    else if (v != "PWGTP") {
      number_of_numeric_values_outside_pwgtp <- number_of_numeric_values_outside_pwgtp + 1
    }
  }
  
  if (non_allowed_nuermic_var_in_input) {
    stop("numeric_vars must only contain: ",
         paste(allowed_numeric_vars, collapse = ", "))
  }
  
  if (number_of_numeric_values_outside_pwgtp == 0) {
    stop("at least one numeric var other than PWGTP must be specified in the input")
  }
  
  if (!is_pwgtp_part_of_input) {
    numeric_vars <- c(numeric_vars, "PWGTP")
  }
  
  
  allowed_categorical_vars <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  non_allowed_categorical_var_in_input <- FALSE
  numb_categorial_variables <- 0
  # validate the input tags
  for (v in categorical_vars) {
    v_allowed <- FALSE
    for (allowed in allowed_categorical_vars) {
      if (v == allowed) {
        v_allowed <- TRUE
        numb_categorial_variables <- numb_categorial_variables + 1
      }
    }
    
    if (!v_allowed) {
      non_allowed_categorical_var_in_input <- TRUE
    }
  }
  
  if (non_allowed_categorical_var_in_input) {
    stop(
      "categorical_vars must only contain: ",
      paste(allowed_categorical_vars, collapse = ", ")
    )
  }
  
  if (numb_categorial_variables == 0) {
    stop(
      "at least one valid categorical must be specified in the input. categorical_vars must only contain: ",
      paste(allowed_categorical_vars, collapse = ", ")
    )
  }
  

   
 for (v in categorical_vars) {
    v_allowed <- FALSE
    for (allowed in allowed_categorical_vars) {
      if (v == allowed) {
        v_allowed <- TRUE
        numb_categorial_variables <- numb_categorial_variables + 1
      }
    }
    
    if (!v_allowed) {
      non_allowed_categorical_var_in_input <- TRUE
    }
  }
  
    #let's validate geography 
  
    variables_metadata <- get_variables_by_year(year_str)

  
geography_level <- match_allowed_value(geography_level, c("Region", "Division", "State"))
if (is.na(geography_level)) {
  stop("geography_level must be one of: Region, Division, State")
}

if (geography_level == "State") {
   if (year_str == "2021" || year_str == "2022") {
    geography_var_name <- "ST"
  } else {
    geography_var_name <- "STATE"
  }
  geography_for_key <- "state"
} else if (geography_level == "Division") {
  geography_var_name <- "DIVISION"
   geography_for_key <- "division"
} else if (geography_level == "Region") {
  geography_var_name <- "REGION"
   geography_for_key <- "region"
} 

print(paste0("geography_var_name,", geography_var_name))


allowed_geography_values <- names(variables_supported[[geography_var_name]]$values$item)
print(allowed_geography_values)

geography_value <- match_allowed_value(geography_value, c(allowed_geography_values, "all"))
if (is.na(geography_value)) {
  stop("geography_value must be 'all' or one of the following valid values for ", geography_level, ": ",
       paste(allowed_geography_values, collapse = ", "))
}


geography_clause <- if (geography_value == "all") {
  paste0("&for=", geography_for_key, ":*")
} else {
  paste0("&for=", geography_for_key, ":", geography_value)
}
  
  url <- paste0(
    "https://api.census.gov/data/",
    year_str,
    "/acs/acs1/pums?get=",
    paste(numeric_vars, collapse = ","),
    ",",
    paste(categorical_vars, collapse = ","),
      ",",
    geography_var_name,
    geography_clause,
    "&SCHL=24&key=",
    api_key
  )
  
  print(url)
  
  pums_data <- GET(url)
  resp_json <- fromJSON(rawToChar(pums_data$content))
  
  pums_tibble <- as_tibble(resp_json[-1, , drop = FALSE]) # without -1 header becomes the first row
  colnames(pums_tibble) <- resp_json[1, ]
  
  
  
  for (col_name in colnames(pums_tibble)) {
    #print(variables_supported[col_name])
    if (col_name %in% c("JWAP", "JWDP")) {
      pums_tibble[[col_name]] <- convert_time_codes_to_midpoints(pums_tibble[[col_name]], variables_supported[[col_name]])
    } else if (col_name %in% allowed_numeric_vars) {
      pums_tibble[[col_name]] <- as.numeric(pums_tibble[[col_name]])
    } else if (col_name %in% allowed_categorical_vars) {
     pums_tibble[[col_name]] <- factor(pums_tibble[[col_name]], levels = names(variables_supported[[col_name]]$values$item))
     levels(pums_tibble[[col_name]]) <- unname(unlist(variables_supported[[col_name]]$values$item)) # labels

     }
  }
  
  
  class(pums_tibble) <- c("census", class(pums_tibble))
  pums_tibble
}


pums_tibble_2024_formatted <- get_pums_data_as_tibble__by_year_formatted_with_categorical_variables(
  api_key,
  c("2024"),
  numeric_vars = c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP"),
  categorical_vars = c("FER","HHL"),
  geography_level = c("division"),
  geography_value = c(2)
)
print(pums_tibble_2024_formatted,n = 200)


```

##Combining Years

```{r}
get_pums_data_as_tibble_multiple_years <- function(api_key,
                                                   year_strs = c("2024"),
                                                   numeric_vars = c("AGEP", "PWGTP"),
                                                   categorical_vars = c("SEX"),
                                                   geography_level = c("state"),
                                                   geography_value = c("37")) {
  allowed_years <- c("2024","2023","2022","2021")
  for(yr in year_strs) {
    yr_allowed <- FALSE
    for(allowed in allowed_years) {
      if(yr == allowed) {
        yr_allowed <- TRUE
      }
    }
    if(!yr_allowed) {
      if (!yr_allowed) {
      stop("year_strs must only contain: ", paste(allowed_years, collapse = ", "))
    }
    }
  }
  
  combined_tibble <- NULL
   for(yr in year_strs) {
     year_tibble <- get_pums_data_as_tibble__by_year_formatted_with_categorical_variables(
      api_key,
      yr,
      numeric_vars = numeric_vars,
      categorical_vars = categorical_vars,
      geography_level = geography_level,
      geography_value = geography_value
    )
     
     #normalize ST to STATE to that combining works
     if ("ST" %in% colnames(year_tibble)) {
    colnames(year_tibble)[colnames(year_tibble) == "ST"] <- "STATE"
  }
     
     year_tibble$year <- yr
     
     if(is.null(combined_tibble)) {
       combined_tibble <- year_tibble
     } else {
       combined_tibble <- dplyr::bind_rows(combined_tibble, year_tibble)
     }
   }
  
  class(combined_tibble) <- c("census", class(combined_tibble))
  combined_tibble
  
}

pums_tible_2024_2021_with_state <- get_pums_data_as_tibble_multiple_years(
  api_key,
  c("2024","2021"),
  numeric_vars = c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP"),
  categorical_vars = c("FER", "HHL", "SCH", "SCHL", "SEX"),
  geography_level = c("state"),
  geography_value = c(37)
)

pums_tible_2024_2021_with_state
print(unique(pums_tible_2024_2021_with_state$year))
table(pums_tible_2024_2021_with_state$year)

```

# Part 2: Writing a Generic Function for Summarizing

```{r}
summary.census <- function(x, numeric_vars = NULL, categorical_vars = NULL) {
  
  dflt_categorical_vars <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  dflt_numeric_vars <- c("AGEP", "GASP", "GRPIP",  "JWMNP")

  if(is.null(numeric_vars)) {
    numeric_vars = dflt_numeric_vars
  } else {
    for(n in numeric_vars) {
      matched_value <- match_allowed_value(n,dflt_numeric_vars)
      if(is.na(matched_value)) {
        stop("numeric variables must only contain: ", paste(dflt_numeric_vars,collapse = ","))
      }
    }
  }
  
  if(is.null(categorical_vars)) {
    categorical_vars = dflt_categorical_vars
  } else {
    for(c in categorical_vars) {
      matched_value <- match_allowed_value(c,dflt_categorical_vars)
      if(is.na(matched_value)) {
        stop("numeric variables must only contain: ", paste(dflt_categorical_vars,collapse = ","))
      }
    }
  }
  
  summary_result <- list()
  
  print(class(numeric_vars))
print(length(numeric_vars))

  
  for(v in numeric_vars) {
    
    print(v)
print(length(v))

    numeric_vector <- x[[v]]
    weight_vector <- x[["PWGTP"]]
    sample_mean <- sum(numeric_vector * weight_vector) / sum(weight_vector)
    sample_sd = sqrt(sum(numeric_vector^2 * weight_vector)/sum(weight_vector) - sample_mean^2)
    
    summary_result[[v]] <- list(mean = sample_mean, sd = sample_sd)
  }
  
  for(c in categorical_vars) {
    summary_result[[c]] <- table(x[[c]])
  }
  
  summary_result

}

```

```{r}
summary(pums_tible_2024_2021_with_state)
```
